From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-30 upi.tex
expand on first part of section 2.1

The examples below assume a representation of the type of booleans
$\bt$ as the disjoint union $\ot \oplus \ot$ with the left injection
representing $\mathsf{false}$ and the right injection representing
$\mathsf{true}$. Given an arbitrary reversible function
$\AgdaFunction{f}$ of type $a \leftrightarrow a$, we can build the
reversible function $\AgdaFunction{controlled}~\AgdaFunction{f}$ that
takes a pair of type $\bt \otimes a$ and checks the incoming boolean;
if it is false (i.e., we are in the left injection), the function
behaves like the identity; otherwise the function applies
$\AgdaFunction{f}$ to the second argument. In either case, the
incoming boolean is reconstituted to maintain reversibility. Using the
\AgdaFunction{controlled} building block, we can build a
controlled-not $\AgdaFunction{cnot}$ gate and a
controlled-controlled-not gate, also known as the
\AgdaFunction{toffoli} gate. The latter gate is a universal function
for combinational boolean circuits thus showing the expressiveness of
the language:

@@

{\small
\[\def\arraystretch{1.2}\begin{array}{rcl}
\AgdaFunction{controlled} &:& \forall a.~ (a \leftrightarrow a) \rightarrow 
                              (\bt \otimes a \leftrightarrow \bt \otimes a) \\
\AgdaFunction{controlled}~\AgdaFunction{f} &=& 
      \bt \otimes a \\
&& \qquad\byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
&& (\ot \oplus \ot) \otimes a \\
&& \qquad\byiso{\AgdaFunction{distribute}} \\
&& (\ot \otimes a) \oplus (\ot \otimes a) \\
&& \qquad\byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{f})} \\
&& (\ot \otimes a) \oplus (\ot \otimes a) \\
&& \qquad\byiso{\AgdaFunction{factor}} \\
&& (\ot \oplus \ot) \otimes a \\
&& \qquad\byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
&& \bt \otimes a \\
\\ 
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=& 
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\ 
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\ 
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
\end{array}\]}
%%%

>>>>

The examples below assume a representation of the type of booleans
$\bt$ as the disjoint union $\ot \oplus \ot$ with the left injection
representing $\mathsf{false}$ and the right injection representing
$\mathsf{true}$. Given an arbitrary reversible function
$\AgdaFunction{f}$ of type $a \leftrightarrow a$, we can build the
reversible function $\AgdaFunction{controlled}~\AgdaFunction{f}$ that
takes a pair of type $\bt \otimes a$ and checks the incoming boolean;
if it is false (i.e., we are in the left injection), the function
behaves like the identity; otherwise the function applies
$\AgdaFunction{f}$ to the second argument. In either case, the
incoming boolean is reconstituted to maintain reversibility.

{\small
\[\def\arraystretch{1.2}\begin{array}{rcll}
\AgdaFunction{controlled}  &:& \forall a.~ (a \leftrightarrow a) \rightarrow 
                            & (\bt \otimes a \leftrightarrow \bt \otimes a) \\
\AgdaFunction{controlled}~\AgdaFunction{f} &=&

  \bt \otimes a 
    & \byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
&& (\ot \oplus \ot) \otimes a 
    & \byiso{\AgdaFunction{distribute}} \\
&& (\ot \otimes a) \oplus (\ot \otimes a) 
    & \byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{f})} \\
&& (\ot \otimes a) \oplus (\ot \otimes a)
    & \byiso{\AgdaFunction{factor}} \\
&& (\ot \oplus \ot) \otimes a
    & \byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
&& \bt \otimes a & ~ \\
\end{array}
\]}

The $\AgdaFunction{not}$ function is a simple lifting of
\AgdaFunction{swap₊}. Using the
\AgdaFunction{controlled} building block, we can build a
controlled-not $\AgdaFunction{cnot}$ gate and a
controlled-controlled-not gate, also known as the
\AgdaFunction{toffoli} gate. The latter gate is a universal function
for combinational boolean circuits thus showing the expressiveness of
the language.

@@

{\small
\[\begin{array}{rcl}
\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
\AgdaFunction{not} &=& 
  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
\\ 
\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
\\ 
\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
\end{array}\]}
%%%

While we wrote \AgdaFunction{controlled} in equation-reasoning style,
\AgdaFunction{not} is written in combinator style.  These are
equivalent as $\byiso{-}$ is defined in terms of $\odot$.
